Sampling Strategies in Decisions from Experience

### 2022-02-19 manuscript.Rmd
Update Section 2 (CPT)

The choices people actually make between prospects---be that in DfD or DfE---are often described in terms of deviations from the principle of mathematical expectation, according to which the prospect with the largest expected value (EV)
$$
EV = \sum_i^n x_i \times p_X(x_i)
\; .
\tag{9}
$$ 
should be chosen (EV maximization).
To *describe* how people's actual choices deviate from this principle [see @erevAnomaliesForecastsDescriptive2017, for a recent replication of classical demonstrations], the models of CPT [@tverskyAdvancesProspectTheory1992] and similar rank-dependent models [see @stottCumulativeProspectTheory2006, for an overview of models] fit choice data by maximizing^[<!--Open footnote-->Note that CPT may be amended by a stochastic choice rule to account for the probabilistic nature of choices [@rieskampProbabilisticNaturePreferential2008]. Most choice rules [@stottCumulativeProspectTheory2006, for an overview] are functions of the prospects' values $V$ and make a choice of a prospect with a greater $V$ only probable but not deterministic.<!--Close footnote-->]
the value
$$
V = \sum_i^n v(x_i) \times \pi_i
\; ,
\tag{10}
$$ 
with objective outcomes $x_i$ being transformed by a *value function* $v$, and *cumulative decision weights* $\pi_i$ being determined by the difference between transformed cumulative probabilities of the distribution $P_X$.
Importantly, the transformation of cumulative probabilities is carried out by a nonlinear weighting function $w$ (see (12) and (13) below).

To break down (9) and (10) and give some further background on the as-if weighting terminology in the description of choices, leaving out mathematical detail and much of the modeling jargon for now:^[<!--Open footnote-->See, @farrellComputationalModelingCognition2018, for an accessible introduction to fundamental modeling concepts.<!--Close footnote-->] 
Following the principle of mathematical expectation, one would calculate the prospects' EVs according to (9) to subsequently choose the prospect with the greatest EV (EV maximization).
However, in quite many occasions, people do not choose the prospect with the greatest EV [e.g., @erevAnomaliesForecastsDescriptive2017; @kahnemanProspectTheoryAnalysis1979].
To describe people's actual choices in a way that effectively captures the deviations from the principle of mathematical expectation,^[<!--Open footnote-->@tverskyAdvancesProspectTheory1992 addressed the *fourfold pattern of risk attitudes*---i.e., choices between prospects indicate risk aversion for gains of high probability and losses of low probability, and risk seeking for gains of low probability and losses of high probability---and related effects, e.g., the *certainty effect* and the *reflection effect*, that were already addressed by CPT's predecessor, prospect theory [@kahnemanProspectTheoryAnalysis1979]. Together, these empirical choice patterns indicate a violation of the principle of mathematical expectation and the expected utility (EU) principle [see @bernoulliExpositionNewTheory1954\/1738].<!--Close footnote-->]
CPT uses a value function $v$ and a weighting function $w$ to estimate how the objective outcomes and probabilities in (9) can be adjusted, i.e., replaced by subjective values $v(x_i)$ and cumulative decision weights $\pi_i$, respectively, so that the maximization of the value in (10) would produce model-implied choice data that is, roughly speaking, similar to the actual choice data.
Essentially, the proposed model is such that the estimated cumulative decision weights in (10) may be greater or smaller than the respective objective probabilities in (9); and they may therefore give more or less weight to the subjective values than would be warranted by the objective probability.
In other words, for a cumulative decision weight that is greater (smaller) than the objective probability it replaces, the respective outcome---which is transformed into a subjective value---is overweighted (underweighted) *in* CPT.
Accordingly, to describe an empirical choice pattern, one may adopt the as-if weighting terminology which was already used in the introduction by stating that people decide *as if* they maximized the value in (10) and applied the over- and underweighting pattern that was estimated in CPT [see @pachurWeightUncertainEvents2019].
<!--Notably, the as-if prefix indicates that it cannot be concluded from the estimated cumulative decision weights that the mind does indeed perform any of the computations associated with the weighting of subjective values in CPT; rather, it processes the inputs in a way that the outputs, i.e., the actual choice patterns, approximate the solutions from the estimated CPT model [see, e.g., @griffithsRationalUseCognitive2015].
Therefore, this paper uses CPT merely as a data model to describe choices between prospects and to demonstrate how different sampling strategies can produce weighting patterns in CPT resembling those that were estimated for empirical choice data from DfE.-->

Now, more formally, following @tverskyAdvancesProspectTheory1992, the objective outcomes are transformed by a value function
$$
\begin{aligned}
  v : \Omega' &\to \mathbb{R} \\
  x &\mapsto 
  \begin{cases}
     x_i^\alpha &\forall x_i \geq 0, \\
     -\lambda \times |x_i|^\alpha &\text{else}
     \; ,
  \end{cases}
\end{aligned}
\tag{11}
$$
with $\alpha \in [0,1]$ determining the degree of the function's concavity (convexity) over the positive (negative) outcome interval, and $\lambda > 1$ increasing the function's slope over the negative outcome interval only.
Each subjective value $v(x)$ is then multiplied (or: weighted) with a cumulative decision weight that takes the form
$$
\pi_i =
  \begin{cases}
     w^+(P(X \geq x_i)) - w^+(P(X > x_i)) \quad \forall x_i \geq 0, \\
     w^-(P(X \leq x_i)) - w^-(P(X < x_i)) \quad \text{else}
     \; , 
  \end{cases}
\tag{12}
$$
where $w$ is a monotonic increasing, nonlinear function $w: [0,1] \to [0,1]$ satisfying $w^+(0) = w^-(0) = 0$ and $w^+(1) = w^-(1) = 1$.
Several consequences for the weighting pattern that can be estimated in CPT follow from (12) and are reviewed one by one next:

@@

* The value of a cumulative decision weight depends on the estimated shape of $w$.
That is, the shape of $w$ determines over which interval on the cumulative probability scale $[0,1]$ the images of $w$ take values that are greater or smaller than the respective cumulative probabilities---i.e., the preimages---and how much greater or smaller these image values are.
Hence, assuming a nonlinear shape of $w$, the same numerical difference between objective cumulative probabilities may translate to cumulative decision weights of different value.

@@

\noindent
Given that the value of each cumulative decision weight depends on the two respective cumulative probabilities in (12)---i.e., the probability of obtaining a positive (negative) outcome equal to or greater (smaller) than a respective outcome $x$, and the probability of obtaining a strictly greater (smaller) outcome---the remaining consequences for the weighting pattern in CPT may be reviewed by considering the actual shape of the weighting function.
Therefore, Figure 1 illustrates some of the possible shapes of the two-parameter weighting function of @goldsteinExpressionTheoryPreference1987, which, however, is just one of several parameterizations that have been proposed [e.g., @prelecProbabilityWeightingFunction1998; @tverskyAdvancesProspectTheory1992; see @stottCumulativeProspectTheory2006, for an overview].
Now, let each $p$ on the abscissa be one of the four cumulative probabilities from (12). 
Then each graph in Figure 1 illustrates the shape of the weighting function
$$
\begin{aligned}
  w : [0,1] &\to [0,1] \\
  p &\mapsto \frac{\delta \times p^{\gamma}}
  {\delta \times p^{\gamma} + (1-p)^{\gamma}}
  \; ,
\end{aligned}
\tag{13}
$$
for the respective values of $\gamma \in [0,2]$ and $\delta > 0$.

In essence, if the estimate of a cumulative decision weight is larger (smaller) than the objective probability from which it is derived, the weighting function runs above (below) the dashed identity line.
The choice data are then described *as if* people maximized the value in (8) and applied the over- and underweighting pattern indicated by the estimated shape of the overall probability weighting function.

(ref:weighting) Possible Shapes of Goldstein and Einhorn's -@goldsteinExpressionTheoryPreference1987 Weighting Function

@@

>>>>

The choices people actually make between prospects---be that in DfD or DfE---are often described in terms of deviations from the principle of mathematical expectation, according to which the prospect with the largest expected value (EV)
$$
EV = \sum_i^n x_i \times p_X(x_i)
\; .
\tag{9}
$$ 
should be chosen (EV maximization).
To *describe* how people's actual choices deviate from this principle [see @erevAnomaliesForecastsDescriptive2017, for a recent replication of classical demonstrations],^[<!--Open footnote-->@tverskyAdvancesProspectTheory1992 addressed the *fourfold pattern of risk attitudes*---i.e., choices between prospects indicate risk aversion for gains of high probability and losses of low probability, and risk seeking for gains of low probability and losses of high probability---and related effects, e.g., the *certainty effect* and the *reflection effect*, that were already addressed by CPT's predecessor, prospect theory [@kahnemanProspectTheoryAnalysis1979]. Together, these empirical choice patterns indicate a violation of the principle of mathematical expectation and the expected utility (EU) principle [see @bernoulliExpositionNewTheory1954\/1738].<!--Close footnote-->]
the models of CPT [@tverskyAdvancesProspectTheory1992] and similar rank-dependent models [see @stottCumulativeProspectTheory2006, for an overview of models] fit choice data by maximizing^[<!--Open footnote-->Note that CPT may be amended by a stochastic choice rule to account for the probabilistic nature of choices [@rieskampProbabilisticNaturePreferential2008]. Most choice rules [@stottCumulativeProspectTheory2006, for an overview] are functions of the prospects' values $V$ and make the choice of a prospect more probable the greater its $V$ relative to those of other prospects. That is, the greatest $V$ does not determine that the respect prospect is chosen by the model.<!--Close footnote-->]
the value
$$
V = \sum_i^n v(x_i) \times \pi_i
\; ,
\tag{10}
$$ 
with objective outcomes $x_i$ being transformed by a *value function* $v$, and *cumulative decision weights* $\pi_i$ being determined by the difference between transformed cumulative probabilities of the distribution $P_X$.
<!--Importantly, the transformation of cumulative probabilities is carried out by a nonlinear weighting function $w$ (see (12) and (13) below).

To break down (9) and (10) and give some further background on the as-if weighting terminology in the description of choices, leaving out mathematical detail and much of the modeling jargon for now:
Following the principle of mathematical expectation, one would calculate the prospects' EVs according to (9) to subsequently choose the prospect with the greatest EV.
However, quite often, people do not choose the prospect with the greatest EV [e.g., @erevAnomaliesForecastsDescriptive2017; @kahnemanProspectTheoryAnalysis1979].
To describe people's actual choices in a way that effectively captures the deviations from the principle of mathematical expectation,^[@tverskyAdvancesProspectTheory1992 addressed the *fourfold pattern of risk attitudes*---i.e., choices between prospects indicate risk aversion for gains of high probability and losses of low probability, and risk seeking for gains of low probability and losses of high probability---and related effects, e.g., the *certainty effect* and the *reflection effect*, that were already addressed by CPT's predecessor, prospect theory [@kahnemanProspectTheoryAnalysis1979]. Together, these empirical choice patterns indicate a violation of the principle of mathematical expectation and the expected utility (EU) principle [see @bernoulliExpositionNewTheory1954\/1738].]
CPT uses a value function $v$ and a weighting function $w$ to estimate how the objective outcomes and probabilities in (9) can be adjusted, i.e., replaced by subjective values $v(x_i)$ and cumulative decision weights $\pi_i$, respectively, so that the maximization of the value in (10) would produce model-implied choice data that are, roughly speaking, similar to the actual choice data.
Essentially, the proposed model is such that the estimated cumulative decision weights in (10) may be greater or smaller than the respective objective probabilities in (9); and they may therefore give more or less weight to the subjective values than would be warranted by the objective probability.
In other words, for a cumulative decision weight that is greater (smaller) than the objective probability it replaces, the respective transformed outcome is overweighted (underweighted) *in* CPT.
Accordingly, to describe empirical choices, one may adopt the as-if weighting terminology which was already used in the introduction by stating that people choose *as if* they maximized the value in (10) and applied the weighting pattern that was estimated in CPT [see @pachurWeightUncertainEvents2019].
Importantly, the exact weighting pattern is captured by CPT's weighting function which, as illustrated in Figure 1, can take rather distinct graphical shapes.
That is, each graph in Figure 1 implies a particular weighting pattern, with some of them being similar and others being rather distinct, as is described below the following paragraph.--> 
More specifically, following @tverskyAdvancesProspectTheory1992, the objective outcomes are transformed by a value function
$$
\begin{aligned}
  v : \Omega' &\to \mathbb{R} \\
  x &\mapsto 
  \begin{cases}
     x_i^\alpha &\forall x_i \geq 0, \\
     -\lambda \times |x_i|^\alpha &\text{else}
     \; ,
  \end{cases}
\end{aligned}
\tag{11}
$$
with $\alpha \in [0,1]$ determining the degree of the function's concavity (convexity) over the positive (negative) outcome interval, and $\lambda > 1$ increasing the function's slope over the negative outcome interval only.
Each subjective value $v(x)$ is then multiplied (or: weighted) with a cumulative decision weight that takes the form
$$
\pi_i =
  \begin{cases}
     w^+(P(X \geq x_i)) - w^+(P(X > x_i)) \quad \forall x_i \geq 0, \\
     w^-(P(X \leq x_i)) - w^-(P(X < x_i)) \quad \text{else}
     \; , 
  \end{cases}
\tag{12}
$$
where $w$ is a monotonic increasing, nonlinear *weighting function* $w: [0,1] \to [0,1]$ satisfying $w^+(0) = w^-(0) = 0$ and $w^+(1) = w^-(1) = 1$.

Several consequences for the weighting pattern that can be estimated in CPT follow from (12) and are reviewed one by one next:

@@

* The value of a cumulative decision weight depends on the estimated shape of the graph of $w$, the function's parameters respectively.
That is, the shape of the graph of $w$ displays over which interval on the cumulative probability scale $[0,1]$ the images of $w$ take values that are greater or smaller than the respective cumulative probabilities---i.e., the preimages---and how much greater or smaller these image values are.
Hence, assuming a nonlinear graphical shape of $w$, the same numerical difference between objective cumulative probabilities may translate to cumulative decision weights of different value.

@@

\noindent
Given that the value of each cumulative decision weight depends on the two respective cumulative probabilities in (12)---i.e., the probability of obtaining a positive (negative) outcome equal to or greater (smaller) than a respective outcome $x$, and the probability of obtaining a strictly greater (smaller) outcome---the remaining consequences for the weighting pattern in CPT may be reviewed by considering the actual weighting function.
Therefore, Figure 1 illustrates some of the possible graphical shapes of the two-parameter weighting function of @goldsteinExpressionTheoryPreference1987, which, however, is just one of several parameterizations that have been proposed [e.g., @prelecProbabilityWeightingFunction1998; @tverskyAdvancesProspectTheory1992; see @stottCumulativeProspectTheory2006, for an overview].
Now, let each $p$ on the abscissa be one of the four cumulative probabilities from (12). 
Then each graph in Figure 1 displays the graphical shape of the weighting function
$$
\begin{aligned}
  w : [0,1] &\to [0,1] \\
  p &\mapsto \frac{\delta \times p^{\gamma}}
  {\delta \times p^{\gamma} + (1-p)^{\gamma}}
  \; ,
\end{aligned}
\tag{13}
$$
for the respective values of the free parameters $\gamma \in [0,2]$ and $\delta > 0$. 
Evidently, both parameters have distinct effects on the graphs' shape, with $\gamma$ affecting its curvature and $\delta$ the elevation.

(ref:weighting) Possible Graphical Shapes of Goldstein and Einhorn's -@goldsteinExpressionTheoryPreference1987 Weighting Function

@@

To begin with, the gray-dashed identity lines in Figure 1 imply a linear weighting pattern.
That is, since the images of $w$, $w(p)$, are equal to the respective cumulative probabilities $p$, all cumulative decision weights $\pi$ that are retrieved according to (12) would be equal to the difference between the respective objective cumulative probabilities and therefore also to the objective probabilities of the outcomes.
Such a linear weighting pattern is implied by $\gamma = 1$ and $\delta = 1$. 
Now, both, deviations from $\gamma = 1$ and $\delta = 1$ produce a nonlinear graphical shape. 
That is, for $\gamma > 1$ (green graphs), the graphs take a S-shape---running below (above) the identity line for small (large) cumulative probabilities---which is more pronounced for larger deviations from $\gamma = 1$.
Importantly, the same *small* numerical difference between two cumulative probabilities $p$ translates to varying differences between the respective images $w(p)$ depending on the interval on the cumulative probability scale. 
That is, because of the S-shaped curvature, the differences between the images $w(p)$ may be smaller than those of the respective cumulative probabilities in the lower and upper part of the cumulative probability scale, but may be considerably greater in the middle part. 
As a consequence, the cumulative decisions weights that are retrieved from the images of $w$ may be smaller than the respective objective probabilities for small-probability outcomes of either a low rank or a high rank, but may be larger for small-probability outcomes of middle ranks. 
Regarding the underweighting of small-probability outcomes of low or high rank, the weighting pattern that is captured by the graphical S-shape of CPT's weighting function is denoted the underweighting of rare outcomes.
Note that for all $\gamma < 1$ (blue graphs) the graph of $w$ takes an inverse S-shape and the entire weighting pattern---denoted as the overweighting of rare outcomes---is reversed.
Nonetheless, the effects of $\gamma$ depend rather strongly on $\delta$, which affects the elevation of the graph.
As can be seen,
